---
title: "Chinook Index V1"
output: pdf_document
---

This is an initial fitting of Sean's spatio-temporal random fields model (`sdmTMB`) to the Chinook data from the High Seas and IPES surveys. I had two main objectives:

1. Determine whether there were sufficient differences in Chinook catch between day/night tows to justify excluding the latter for historical comparisons.
2. Evaluate how including two different subsets of the high seas data impacts predicted estimates of annual abundance.


```{r load_clean}
library(tidyverse)
library(sdmTMB)
library(ggplot2)
library(glmmTMB)

# browseVignettes("sdmTMB")

bridge <- readRDS(here::here("data", "ipes_hs_merged_bridge.rds")) 

# use bridge data alone for now since ignoring stock composition
jchin <- bridge %>%
  #scale UTM coords
  mutate(xUTM_start = xUTM_start / 10000,
         yUTM_start = yUTM_start / 10000,
         yday_z = as.vector(scale(yday)[,1]),
         yday_z2 = yday_z^2,
         time_f = as.factor(time_f),
         juv_cpue = ck_juv / dur
         ) %>% 
  #remove extra vars and stock ppn data
  dplyr::select(station_id, stable_station:date, time_f, yday, yday_z, yday_z2,
                month:dur, head_depth,
                ck_juv, juv_cpue)
```

** Day/Night Comparison **

To evaluate day/night impacts I fit a zero-inflated negative binomial model to the IPES data. The day/night intercept was the only fixed effect and year was treated as a random intercept to account for covariance in abundance within a survey. Given that the survey is intentionally structured to be randomly stratified I believe additional covariates are probably unnecessary for this particular question.

```{r day_night_fit}
ipes_only <- jchin %>% 
  filter(synoptic == "1",
         month %in% c(6, 7),
         dataset == "IPES")

fit_zibin <- glmmTMB(ck_juv ~ time_f + (1|year), data = ipes_only, 
                     ziformula = ~1, family = nbinom2)
summary(fit_zibin)
```


Parameter estimates suggest there is little evidence that Chinook abundance differs between sampling periods; however, interpreting coefficients on the log scale is difficult so we generate predictions to see what model suggests the mean abundance (all else being equal) from a day vs night tow would be.


```{r day_night_predict}
# make predictions
preds <- predict(fit_zibin,
        newdata = data.frame(time_f = unique(ipes_only$time_f),
                             year = NA),
        se.fit = TRUE,
        #generates predictions on scale of resp. rather than link function
        type = "response") 

# plot predictions
data.frame(time_f = unique(ipes_only$time_f),
                       fit = preds$fit,
                       se = preds$se.fit) %>% 
  mutate(low = fit + qnorm(0.025) * se,
         up = fit + qnorm(0.975) * se) %>% 
  ggplot(.) +
  geom_pointrange(aes(x = time_f, y = fit, ymin = low, ymax = up)) +
  labs(x = "Sampling Time", y = "Predicted Abundance Per Tow") +
  ggsidekick::theme_sleek()
```


So the night data predict an abundance of ~0.35 and day of 0.42 juvenile Chinook per tow, but there's substantial overlap in the confidence intervals. I also tried fitting similar models with tweedie and poisson distributions. They generated similar results, but weren't supported given AIC scores. Unless we identify other covariates that might confound this relationship I would suggest that, at least for juvenile Chinook, we can pool day and night tows when making annual indices.


** Annual Index **

First generate two datasets that subset the high seas data in different ways (the goal is to retain as much as is appropriate since a data accouting for stock composition will be very data hungry). Both datasets are constrained to only include June/July catches (though I'd like to eventually look into incorporating more data) and tows less than 21 m deep. The less conservative dataset includes all the regions where spatial overlap of the two surveys occurs (e.g. WCVI inlets, QCS). The second includes only the high seas surveys that fall within the IPES standard survey grid (based on Erika's SQL code).

```{r annual_index_data}
jchin1 <- jchin %>% 
  filter(stable_station == "1",
         month %in% c(6, 7),
         head_depth < 21,
         #remove some exceptionally large tows that produce wacky residuals
         !juv_cpue > 500 
         ) %>% 
  mutate(trim_size = "large") 
jchin2 <- jchin %>% 
  filter(synoptic == "1",
         head_depth < 21,
         month %in% c(6, 7),
         !year == "2014"
         ) %>% 
  mutate(trim_size = "small") 

# compare spatial distributions of the two subsets
library(ggmap)
nAm <- map_data("world") %>% 
  filter(region %in% c("Canada", "USA"))
rbind(jchin1, jchin2) %>% 
  ggplot(.) +
  geom_point(aes(x = start_long, y = start_lat, color = dataset), alpha = 0.4) +
  geom_map(data = nAm, map = nAm, aes(long, lat, map_id = region),
           color = "black", fill = "gray80") +
  coord_fixed(xlim = c(-129.5, -123), ylim = c(48, 52), ratio = 1.3) +
  facet_wrap(~trim_size) +
  ggsidekick::theme_sleek()

rbind(jchin1, jchin2) %>% 
  ggplot(., aes(x = yday, y = juv_cpue, colour = as.factor(dataset))) +
    geom_point(alpha = 0.5) +
    facet_wrap(~trim_size) +
    ggsidekick::theme_sleek()
```

The majority of both datasets are shared and come from the core IPES sets, so it's unsurprising that they both show similar seasonal patterns in catch (i.e. peaking in late June); howerver, there are certainly some differences in abundance driven by the inlets (presumably more catch) and offshore tows (presumably less catch). We could chop the former dataset a bit more if we're interested in keeping more data but excluding the offshore transects, but I wanted to maximize contrast on this first pass. 

Now we prep the SPDE mesh necessary to account for spatial heterogeneity and fit Sean's models. To account for those seasonal effects we include z-scored year day as a covariate. To account for differences in tow length (the average high seas tow was substantially longer), I fit the models to catch and included tow duration as an offset. The data are not log transformed prior because the negative binomial is intended for positive integer data and can account for a relatively large number of zeros. Note that the run time for these models can be pretty long so I've included the RDS files that you can simply load once you move them to the appropriate location.

```{r annual_index_fit}
# Make mesh representing survey area for each dataset
jchin1_spde <- make_spde(jchin1$xUTM_start, jchin1$yUTM_start, n_knots = 150)
jchin2_spde <- make_spde(jchin2$xUTM_start, jchin2$yUTM_start, n_knots = 150)
#plot_spde(jchin_spde1)
#plot_spde(jchin_spde2)

# import data (note the file path will obviously need to be changed)
m1_nb <- readRDS(here::here("data", "modelFits", "ck1_yrInt_nb.rds"))
m2_nb <- readRDS(here::here("data", "modelFits", "ck2_yrInt_nb.rds"))

# alternatively run models (note tweedie replaced with NB)
# m1_nb <- sdmTMB(ck_juv ~ 0 + as.factor(year) + yday_z + yday_z2 + offset,
#              data = jchin1,
#              time = "year",
#              spde = jchin1_spde,
#              silent = FALSE,
#              anisotropy = TRUE,
#              include_spatial = TRUE,
#              ar1_fields = FALSE,
#              family = nbinom2(link = "log"))
# m2_nb <- sdmTMB(ck_juv ~ 0 + as.factor(year) + yday_z + yday_z2 + offset,
#                 data = jchin2,
#                 time = "year",
#                 spde = jchin2_spde,
#                 silent = FALSE,
#                 anisotropy = TRUE,
#                 include_spatial = TRUE,
#                 ar1_fields = FALSE,
#                 family = nbinom2(link = "log"))
```

We can examine the coefficient estimates using `summary()`, but for now we'll just jump right into predictions. My understanding from Sean is that the predicted fixed effects are generated after accounting spatial variation that's treated as a random field. So differences in specific survey locations among years are integrated out and incorporated as uncertainty in the fixed effect predictions (similar to when you make predictions with a mixed-effects model). The estiamted abundance at each point in the predictive spatial grid are then summed, resulting in an annual index of total abundance for the survey area/time.

I generated the predictive grids externally because they require a bit of code, but let me know if you'd like to see what's happening. Basically for now you can just think of it as making a prediction for a given year, in a given grid cell, assuming scaled yday is set to 0 (i.e. abundance on the "average" summer survey day).

```{r make_preds}
# predictive spatial grids (different for each dataset due to different spatial
# extents)
pred_grid1 <- readRDS(here::here("data", "pred_grids", "large_ck_grid.rds"))
pred_grid2 <- readRDS(here::here("data", "pred_grids", "small_ck_grid.rds"))

pred_m1 <- predict(m1_nb, newdata = pred_grid1, return_tmb_object = TRUE)
pred_m2 <- predict(m2_nb, newdata = pred_grid2, return_tmb_object = TRUE)

# Compare predictions from the two datasets
ind1 <- get_index(pred_m1, bias_correct = FALSE) %>% 
  mutate(dataset = "large")
ind2 <- get_index(pred_m2, bias_correct = FALSE) %>% 
  #add gap for 2014
  rbind(., data.frame(year = 2014, est = NA, lwr = NA, upr = NA, log_est = NA,
                      se = NA, max_gradient = NA, bad_eig = FALSE)) %>% 
  mutate(dataset = "small")  
inds <- rbind(ind1, ind2)

scale <- 2 * 2  # 2 x 2 km grid 
ggplot(inds, aes(year, est*scale)) + 
  geom_line(aes(group = 1)) +
  geom_ribbon(aes(ymin = lwr*scale, ymax = upr*scale), alpha = 0.4) +
  xlab('Year') + ylab('Abundance Estimate') +
  facet_wrap(~dataset)
```

So what can we glean from this? 

1. The two datasets have some similar results including periods of low abundance in the early 2000s and a peak in 2008.
2. Generally the uncertainty is similar, however little survey effort and low catches (i.e. 0) within the IPES grid in years like 2014 means that the model can't converge until they're removed.  
3. Perhaps most importantly the estimates of abundance for recent years are similar, suggesting that the inclusion of additional data in historical years doesn't dramatically change our conclusions about absolute abundance in the IPES survey (it will change relative abundance estimates because large positive anomalies in the high seas database are muted).

This is by no means a final product and will change a lot as I roll in the multinomial component of the model to generate stock-specific predictions, but hopefully it provides some context for what I've been working on recently. 
